---
title: "dplyr + tidyverse"
subtitle: "Session 3 · MLDS Bootcamp · Fall 2026"
author: "Alice Zhao"
engine: knitr
format:
  revealjs:
    theme: white
    transition: fade
    incremental: true
    loop: true
    echo: true
---

## Reshuffle #1

- Find a new group of 2-3 people
- Introduce yourselves
- What did we cover yesterday?

## Agenda

::: nonincremental
- tidyverse & tibbles
- `dplyr`: core verbs
- `tidyr`: reshaping
- `lubridate`: working with dates
:::

## Session 3 Mindset

- We've been using **base R** for data wrangling
- Now we introduce **the tidyverse**: a collection of packages built for data science
- Why tidyverse?
  - More readable, consistent syntax across packages
  - The **pipe** (`%>%` or `|>`) chains steps together
  - Designed specifically for data analysis workflows

## Session 3 Setup

- Open an R script or notebook
- Set your working directory to your Desktop
- Install and load the tidyverse

::: {.fragment}
```{r}
#| eval: false
install.packages("tidyverse")
library(tidyverse)
```
:::

- One package with many packages:
  - `dplyr`, `ggplot2`, `readr`, and more

## Tibbles

- A **tibble** is the tidyverse's version of a data frame
  - Inspired by `tbl_df`
- Behaves like a data frame, with a few upgrades:
  - Nicer printing
  - No partial matching
  - Never changes variable names or types on you

- Create one with `tibble()` or convert with `as_tibble()`

## Data Frames vs Tibbles

```{r}
library(tibble)
df <- data.frame(x = 1:3, y = c("a", "b", "c"))
tb <- tibble(x = 1:3, y = c("a", "b", "c"))
```

```{r}
df
tb
```

## Data Frames vs Tibbles

Example: partial matching

```{r}
#| eval: false
df$x
df$xyz

tb$x
tb$xyz
```

Example: data types

```{r}
#| eval: false
class(df[, "x"])
class(tb[, "x"])
```

## EXERCISE: Tibbles

Read in the `gapminder.csv` as a tibble

::: {.fragment}
Option 1:

```{r}
library(tidyverse)
df <- read.csv('gapminder.csv')
df <- as_tibble(df)
head(df, 3)
```
:::

::: {.fragment}
Option 2: Click on file >> Import Dataset >> Rename as `df`
:::
- Remove environment variables using `rm`

## Tidyverse History

- 1991 (35 years ago): R was created by researchers at University of Auckland in New Zealand
- 2005 (21 years ago): Hadley Wickham came into the scene by creating data science tools in R
  - `ggplot2`: grammar of graphics (formerly `ggplot`)
  - `dplyr`: grammar of data manipulation (formerly `plyr`)
- 2016 (10 years ago): Wickham announced the `tidyverse`, which helped bring multiple libraries together
  
## dplyr Overview

- `dplyr` stands for dataframe pliers
  - Its core verbs are inspired by SQL clauses (`select`, etc.)
  - Additional functions are often updated or added to the library every few years
  
- `dplyr` uses the pipe syntax (`%>%`) from the `magrittr` library
  - Inspired by Magritte's painting *"ceci n'est pas une pipe"*
  - In 2021, base R introduced the `|>` pipe syntax
  
- These are `dplyr's` two core features -- verbs and pipes

## select()

- `select()` picks columns by name
- `df |> select(col1, col2, col3)`
- Use `-` to drop columns instead: `df |> select(-col1)`

## EXERCISE: select()

::: nonincremental
- Select just `country`, `year`, and `gdpPercap` from `df`
- Now select everything *except* `continent`

```{r}
head(df)
```
:::

## SOLUTION: select()

Read as *"df then select country, year and GDP"*

```{r}
df |> select(country, year, gdpPercap)
```

## SOLUTION: select()

Read as *"df then select everything except continent"*

```{r}
df |> select(-continent)
```

## filter()

- `filter()` keeps rows that meet a condition
- `df |> filter(continent == "Asia")`
- Combine conditions with `&` (and) or `|` (or)

## EXERCISE: filter()

::: nonincremental
- Filter `df` to just the year 2007
- Filter to Asia *and* year 2007
:::

## SOLUTION: filter()

```{r}
df |> filter(year == 2007)
```

## SOLUTION: filter()

```{r}
df |> filter(continent == "Asia" & year == 2007)
```

## EXERCISE: Chaining Verbs

::: nonincremental
- Take your dataframe
  - Then select country, year and GDP
  - Then filter to only keep 2007
  - Then only show the top 6 rows
:::

## SOLUTION: Chaining Verbs

```{r}
df |>
  select(country, year, gdpPercap) |>
  filter(year == 2007) |>
  head()
```

::: {style="font-size: 0.9em"}
- Languages like SQL require clauses to be in a fixed written order
- `dplyr` has no required verb order — but there is a conventional one
:::

## SOLUTION: Chaining Verbs

```{r}
df |>
  filter(year == 2007) |> # first reduce the number of rows
  select(country, year, gdpPercap) |> # then reduce the number of columns
  head() # this is almost always at the end
```

## mutate()

- `mutate()` adds or changes a column
  - `df |> mutate(col3 = col1 * col2)`
  - `df |> mutate(col1 = round(col1))`
- Everything in `dplyr` executes sequentially
  - So you can reference a column you just created within the same `mutate()` call

## EXERCISE: mutate()

::: nonincremental
- Create a new column `gdp` equal to `gdpPercap * pop`
- Update `lifeExp` to a rounded `lifeExp`
- Create a new column `gdp_updated` equal to `gdp * 1.02` (*a column you just created within this mutate call*)
:::

## SOLUTION: mutate()

```{r style="font-size: 0.9em"}
df |>
  mutate(
    gdp = gdpPercap * pop,
    lifeExp = round(lifeExp),
    gdp_updated = gdp * 1.02
  )
```


## arrange()

- `arrange()` sorts rows by one or more columns
- Ascending by default
  - `df |> arrange(col1)`
- Wrap a column in `desc()` to flip it
  - `df |> arrange(desc(col1))`

## EXERCISE: arrange()

::: nonincremental
- Arrange `df` by `lifeExp`, from highest to lowest
:::

## SOLUTION: arrange()

```{r}
df |> arrange(desc(lifeExp))
```

## EXERCISE: Saving the Output

- Take your dataframe
  - Then filter to only keep 2007
  - Then select all columns except for `gdpPercap`
  - Then round the `lifeExp` column
  - Then sort the `year` from lowest to highest, and `pop` from highest to lowest
  - Then only show the top 10 rows
- Save the output as `df_filtered`
  
## SOLUTION: Saving the Output

```{r style="font-size: 0.9em"}
df_filtered <- df |>
  filter(year == 2007) |>  # first reduce your data
  select(-gdpPercap) |>
  mutate(lifeExp = round(lifeExp)) |>  # then make calculations
  arrange(year, desc(pop)) |>  # then sort your data
  head(n = 10)  # then display

df_filtered
```

## BREAK! {.center}

## Reshuffle #2

::: nonincremental
- Find a new group of 2-3 people
- Introduce yourselves
:::

## summarize()

- `summarize()` collapses a column down to a single summary value
- `df |> summarize(avg_col1 = mean(col1))`
- Only returns the summary — everything else is dropped

## EXERCISE: summarize()

::: nonincremental
- Find the mean and median `population` across the whole dataset
:::

## SOLUTION: summarize()

```{r}
df |> summarize(avg_pop = mean(pop),
                med_pop = median(pop))
```

## group_by() + summarize()

- `group_by()` splits your data into groups behind the scenes
- `summarize()` then runs separately *for each group*
- `df |> group_by(col1) |> summarize(avg_col2 = mean(col2))`

## EXERCISE: group_by() + summarize()

::: nonincremental
- Find the mean life expectancy in 2007 for each continent (*like Session 2's `aggregate()` exercise, but with `dplyr`*)
- Also find the number of countries in each continent
:::

## SOLUTION: group_by() + summarize()

```{r}
df |>
  filter(year == 2007) |>
  group_by(continent) |>
  summarize(avg_life = mean(lifeExp),
            n_countries = n())
```

## dplyr Verb Summary

- We've now covered the 7 main verbs in `dplyr`
- **EXERCISE**: What is the suggested order of these verbs?
  - arrange
  - filter
  - group_by
  - head
  - mutate
  - select
  - summarize

## SOLUTION: dplyr Verb Summary

```{r}
df |>
  filter(year == 2007) |>              # 1. reduce rows first
  select(-gdpPercap) |>                # 2. drop unneeded columns
  mutate(lifeExp = round(lifeExp)) |>  # 3. transform before grouping
  group_by(continent) |>               # 4. group, if needed
  summarize(avg_le = mean(lifeExp)) |> # 5. collapse to one row per group
  arrange(desc(avg_le)) |>             # 6. sort once data is ready
  head()                               # 7. trim to final rows last
```

## More dplyr Functions

- `dplyr` has many more functions beyond verbs and pipes
  - `count`, `distinct`, etc. -- basic calculations
  - `case_when` -- conditional logic
  - `.by` -- newer, simpler group by syntax created in 2023
  - `join` -- we'll be covering in the next session
  - ... and much more
- Just memorize the core `dplyr` verb and pipe syntax
  - For all other data manipulation tasks -- know what's possible and have AI help with syntax

## Note: Functions with the Same Name

- Some function names exist in multiple packages (e.g. `filter()`)
- R uses whichever package was loaded *most recently*
- Use `package::function()` to be explicit: `dplyr::filter()`

## tidyr: Reshaping Data

- Data can be **wide** or **long**
  - Wide data is more human readable
  - Long data is often needed for data visualization
- `pivot_longer()`: wide → long
- `pivot_wider()`: long → wide

## EXAMPLE: Pivoting

```{r style="font-size: 0.9em"}
relig_income
```

## EXAMPLE: Pivoting

```{r style="font-size: 0.9em"}
relig_income |>
  pivot_longer(
    cols = -religion,
    names_to = "income",
    values_to = "count"
  )
```
- This was an example of going from wide data to long data

## EXERCISE: Pivoting

::: {style="font-size: 0.7em"}

How would you update your table from this:
```{r}
df_before <- tibble(
  country = c("Afghanistan", "Afghanistan", "Afghanistan", "Albania", "Albania", "Albania"),
  year    = c(1997, 2002, 2007, 1997, 2002, 2007),
  lifeExp = c(41.8, 42.1, 43.8, 72.9, 75.6, 76.4)
)
df_before
```

to this?
```{r}
#| echo: false
df_after <- df_before |>
  pivot_wider(names_from = year, values_from = lifeExp)
```

```{r}
df_after
```

:::

## SOLUTION: Pivoting

```{r}
df_after <- df_before |>
  pivot_wider(names_from = year, values_from = lifeExp)
df_after
```

## lubridate: Working with Dates

- Not created by Wickham, but he advised the project
- Dates are notoriously hard to work with in base R
- `lubridate` makes dates and times easier to parse and manipulate
  - `ymd()`, `mdy()`, `dmy()`, etc. parse dates from text
  - `year()`, `month()`, `day()`, etc. extracts the date parts
  - Includes many other datetime capabilities like adding / subtracting dates, specifying time intervals, etc.

## lubridate: Working with Dates

Without `lubridate`, this will return an error:
```{r eval=FALSE}
detach("package:lubridate", unload = TRUE)
year("09/15/2026")
```

<br>

With `lubridate`, the character date can be easily parsed:
```{r}
library(lubridate)
mdy("09/15/2026")
year(mdy("09/15/2026"))
```

## EXERCISE: Working with Dates

Extract the day values from this vector

```{r}
dates <- c("2026-09-14", "2026-09-15",
           "2026-09-16", "2026-09-17",
           "2026-09-18", "2026-09-19")
```

## SOLUTION: Working with Dates

```{r}
library(lubridate)
dates
day(ymd(dates))
day(dates)
```

## Tidyverse Review

What do each of these tidyverse packages help us do?

::: {.fragment .nonincremental}
- dplyr
- tidyr
- lubridate
- tibble
- readr
- ggplot2
:::

## Tidyverse Review

:::: {style="font-size: 0.9em"}

::: {.fragment .nonincremental}
- `dplyr` — manipulate data
- `tidyr` — reshape/tidy data
- `lubridate` — work with dates and times
- `tibble` — the modern dataframe format
- `readr` — read data in and write to files
- `ggplot2` — visualize data
:::

::: {.fragment .nonincremental}
- `stringr` — work with strings/text
- `forcats` — work with factors (categorical variables)
- `purrr` — functional programming / iteration
:::

::::

## CLASS ACTIVITY! {.center}

```{=html}
<style>
.small-text {
  font-size: 0.75em;
}

.tight-break {
  line-height: 0.5;
}

.reveal pre {
  margin-top: 10px;
  margin-bottom: 10px;
}

.reveal .column.pad-left {
  padding-left: 50px;
  box-sizing: border-box;
}
</style>
```
