---
title: "R Project Workflow"
subtitle: "Session 4 · MLDS Bootcamp · Fall 2026"
author: "Alice Zhao"
engine: knitr
format:
  revealjs:
    theme: white
    transition: fade
    incremental: true
    loop: true
    echo: true
---

## Discussion

- What have we covered so far during bootcamp?

## Session 4 Mindset

- Today isn't about one package — it's about **workflow**
  - You already know how to write R code
  - The bigger question is — **how do you organize a real R project**?
- No reshuffle this morning
  - We're building one project from start to finish

## Agenda

::: nonincremental
- See the final project
- R project structure
- Cleaning data
- Transforming data
- Writing functions
- Building plots
- Knitting a report
:::

## DEMO: The Finished Project

- Before we build anything, let's see where we're headed

::: {.fragment .nonincremental}
- Download the `demo-project.zip` file and unzip it
- Open `demo-project.Rproj` and view the contents
- In the **Terminal in R**, run `Rscript run_all.R`
- View `report.html` in a web browser
:::

::: {.fragment .nonincremental}
- Now let's build this from scratch!
:::

## Note: Console vs Terminal

- The **Console** runs R code
- The **Terminal** is like opening the Terminal / cmd app on your machine
  - `Rscript run_all.R` is a shell command
  
## Our Goal Today

- See if there's a relationship between life expectancy and how far a country is from the equator

- The data sets we'll be using are:
  - `gapminder.csv`: includes country and life expectancy data
  - `country_attributes.csv`: includes country and latitude data
  
- Our final deliverable will be to generate a report with a plot and takeaway

## R Project Structure

```{.no-scroll}
demo-project/
├── demo-project.Rproj
├── data/
│   ├── raw/
│   └── processed/
├── R/
│   └── functions.R
├── scripts/
│   ├── 1_clean.R
│   ├── 2_transform.R
│   └── 3_plot.R
├── report.Rmd
└── run_all.R
```

## .Rproj

- This file gets created when you start a new R project
- What is an R project?
  - A folder with R code for a particular app or analysis project that is self-contained (can share with anyone)
- Advantages
  - It automatically sets your working directory so no `setwd()` is needed
  - Environment variables and functions can be shared across scripts
  
## DEMO: Start an R Project

::: {.nonincremental}
- Click New Project button >> New Directory >> New Project
- Name: `my-project`
- Subdirectory: `~/Desktop`
:::

## data

- The data folder contains all the data for your project
- `raw/` is never touched by hand — treat it as read-only
- `processed/` starts **empty** — and then everything in it is generated by an R script

## DEMO: Create a data folder

::: {.nonincremental}
- Create a `data` folder in your `my-project` folder
- Create a subfolder called `raw`
  - Copy over the `gapminder.csv` and `country_attributes.csv` files
- Create a subfolder called `processed`
:::

## scripts

- The scripts folder contains all the R scripts for your project
- They are often numbered so the run order is visible:
  - `1_clean.R`
  - `2_transform.R`
  - `3_plot.R`
- Why are they separated out into many scripts?
  - Each one does exactly one job, which makes it easier to debug and maintain

## Step 1: Data Cleaning

- Let's start by lightly cleaning our data
  - We know that `gapminder.csv` is pretty clean
  - In this script, we'll be cleaning the data in `country_attributes.csv`

## EXERCISE: Read in csv file

::: {.nonincremental}
- Create a `scripts` folder
- Create a new R script inside of it called `1_clean.R`
- Within the script, write code to read in the `country_attributes.csv` file using the tidyverse
:::

## SOLUTION: Read in csv file

```{r}
library(tidyverse)

raw <- read_csv("data/raw/country_attributes.csv")

head(raw)
```

## EXERCISE: Inspect the Raw Data

::: {.nonincremental}
- Before cleaning anything, always look first
- Use `glimpse()`, `duplicated()`, and `n_distinct()` to answer:
  - How many rows and columns?
  - Any duplicate rows?
:::

## SOLUTION: Inspect the Raw Data

```{r style="font-size:0.9em"}
glimpse(raw)
sum(duplicated(raw))
n_distinct(raw$country)
```

## Step 1a: Fix Whitespace

- `" Chile"` and `"Japan "` have leading and trailing spaces
  - They are currently invisible but will break future joins
- `str_trim()` removes them
- `str_to_title()` fixes stray casing like `"germany"` or `"FRANCE"`

## EXERCISE: Trim & Standardize Case

::: {.nonincremental}
- Within your R script, create a new dataframe (tibble) called `formatted`
  - Trim the whitespace in the `country` column
  - Then title-case it
  
<br>

- **Hint:** Remember `dplyr`!
:::

## SOLUTION: Trim & Standardize Case

```{r}
formatted <- raw |>
  mutate(
    country = str_trim(country),
    country = str_to_title(country)
  )
```

## Step 1b: Fix Typos & Alt Names

- Whitespace / case fixes are mechanical — no knowledge is required
- This part isn't: `"Usa"` and `"Brasil"` need someone who knows these should be `"United States"` and `"Brazil"`
- `case_when()` is the function for one-to-one substitutions

<br>

- **Question**: How would you find these errors?
  - Things break down the line!

## EXERCISE: Map the Known Typos

::: {.nonincremental}
- Within your R script, create a new dataframe called `clean_names`
  - map `"Usa"` to `"United States"`
  - map `"Brasil"` to `"Brazil"`
  - leave everything else unchanged
  
<br>

- **Hint**: look up `?case_when` to see syntax examples
:::

## SOLUTION: Map the Known Typos

```{r}
clean_names <- formatted |>
  mutate(
    country = case_when(
      country == "Usa"    ~ "United States",
      country == "Brasil" ~ "Brazil",
      .default = country
    )
  )
```

## Step 1c: Duplicates

- Duplicates should be removed to avoid double counting values
- Use the `distinct()` function to remove duplicate values

::: {.fragment}
```{r}
clean_names |> filter(country == "Canada")
```
:::

## EXERCISE: Remove Duplicates

::: {.nonincremental}
- Within your R script, create a final `cleaned` dataframe that has no duplicate rows
- Save the `cleaned` dataframe inside the `data/processed` folder as `country_attributes_clean.csv`
:::

## SOLUTION: Remove Duplicates

```{r}
cleaned <- clean_names |> distinct()

write_csv(cleaned, "data/processed/country_attributes_clean.csv")
```

<br>

- That's all for `1_clean.R`
- Save it, run it, and check `data/processed/` for the new file

## Steps

1. Clean data ✅
2. Transform data
3. Plot / analyze data

- **Data cleaning** is about fixing messy or inconsistent data
- **Data transformation** is about restructuring data so it's in a more useful format for analysis

## Step 2: Data Transformation

- Remember our goal is to find the relationship between:
  - Life expectancy (from gapminder)
  - Distance from the equator (from country_attributes)
- What columns will we need in our final data set?
  - country
  - lifeExp
  - latitude
- To do this, we're going to need to combine our two tables

## Our current data

We have two clean files:

```{r style="font-size:0.9em"}
gapminder <- read_csv("data/raw/gapminder.csv")
attr_clean <- read_csv("data/processed/country_attributes_clean.csv")
```

```{r style="font-size:0.9em"}
head(gapminder, 3)
head(attr_clean, 3)
```

## Our current data

- We want the `country`, `lifeExp` and `latitude` columns
- Things to notice:
  - Both tables have a `country` **column** — that's our join key
  - The attributes table has a **single country per row**
  - The gapminder table has **multiple rows** for the different years — this needs to be filtered before joining

## EXERCISE: Filter

::: {.nonincremental}
- Create a new file in the `scripts` folder called `2_transform.R`
- Within the file, write code to read in both files (`gapminder.csv`, `country_attributes_clean.csv`)
- Filter the `gapminder` data to `year == 2007`
- Count the number of rows in both dataframes
:::

## SOLUTION: Filter

```{r}
library(tidyverse)

gapminder <- read_csv("data/raw/gapminder.csv")
attr_clean <- read_csv("data/processed/country_attributes_clean.csv")

gapminder_2007 <- gapminder |>
  filter(year == 2007)
```

## SOLUTION: Filter

```{r style="font-size:0.9em"}
nrow(gapminder_2007)
head(gapminder_2007, 2)

nrow(attr_clean)
head(attr_clean, 2)
```

## EXERCISE: Join

- There are 4 main join types (inner, outer, left, right)
  - The one we'll be using in this exercise is the **left join**, which keeps all the countries in the left table

<br>

- **YOUR TASK**
  - Join the tables together using `left_join()` to combine the tables on the common `country` column

## SOLUTION: Join

```{r style="font-size:0.9em"}
joined <- gapminder_2007 |>
  left_join(attr_clean, by = "country")

nrow(joined)
head(joined, 10)
```

## BREAK {.center}

## EXERCISE: Add a New Column

::: {.nonincremental}
- Using the `joined` table, add a column called `dist_from_equator`
  - Distance from the equator = the absolute value of `latitude`
:::

## SOLUTION: Add Derived Columns

```{r}
joined <- joined |>
  mutate(
    dist_from_equator = abs(latitude))

head(joined |> select(country, lifeExp, latitude, dist_from_equator))
```

## DEMO: Add a More Complex Column

- Create an `R` folder and copy the `functions.R` script into it
- Review and understand the logic
- Click `Source` and notice the custom function appear in the Environment
  - This is now ready to be used by other R scripts within this project

## EXERCISE: Add a More Complex Column

::: {.nonincremental}
- Update your most recent `joined` code block to create two columns:
  - `dist_from_equator`: absolute value of `latitude`
  - `equator_zone`: use the `classify_equator_zone` function
- Write the final `joined` dataframe to a file called `final_gapminder_country_attributes_clean.csv` in the `/data/processed` folder
:::

## SOLUTION: Add More Complex Column

```{r}
#| eval: FALSE
joined <- joined |>
  mutate(
    dist_from_equator = abs(latitude),
    equator_zone = classify_equator_zone(dist_from_equator))

write_csv(joined, "data/processed/final_gapminder_country_attributes_clean.csv")
```

<br>

- That's all for `2_transform.R`
- Save it, run it, and check `data/processed/` for the new file

## Steps

1. Clean data ✅
2. Transform data ✅
3. Plot / analyze data

## EXERCISE: Plot the data

::: {.nonincremental}
- Create a new file called `3_plot.R` in the `scripts` folder
- Within the script, write code to plot `dist_from_equator` (x) against `lifeExp` (y)
- Add the takeaway from the plot as the title
:::

## SOLUTION: Plot the data 

```{r}
plot(
  joined$dist_from_equator, joined$lifeExp,
  xlab = "Distance from equator (degrees latitude)",
  ylab = "Life expectancy (years)",
  main = "Life Expectancy Is Higher Farther from the Equator"
)
```

- That's all for `3_plot.R`
- We have now completed all of our R scripts!

## Steps

1. Clean data ✅
2. Transform data ✅
3. Plot / analyze data ✅

## What else is left?

```{.no-scroll}
my-project/
├── my-project.Rproj
├── data/
│   ├── raw/
│   └── processed/
├── R/
│   └── functions.R
├── scripts/
│   ├── 1_clean.R
│   ├── 2_transform.R
│   └── 3_plots.R
├── report_simple.Rmd
└── run_all.R
```

## What else is left?

- You've locked in your data and your code
- Next, let's create:
  1. An R notebook to document our findings
  2. An R script to run all the code
  
## DEMO: Create an R notebook

- Paste in the `report_simple.Rmd` notebook
- View the code and `knit` it together to view the report
  
## DEMO: Run all the scripts

Create a file called `run_all.R`, paste in the following code and save the file:

```{r}
#| eval: FALSE
source("R/functions.R")
source("scripts/1_clean.R")
source("scripts/2_transform.R")
source("scripts/3_plot.R")
rmarkdown::render("report_simple.Rmd")
```

## FINAL STEP! Run it for real

- Save everything and close RStudio

<br>

- Imagine someone shared the `my-project` folder with you
  - Now you can just go to the project folder, and double click on `my-project.Rproj` to open the project
  - Then type `Rscript run_all.R` in the Terminal in R to run all the R scripts and see the outputs

## DISCUSSION

- You just completed your first R project! 🎉
- With a partner, review the steps we went through today
  
## R Project Steps

::: {.nonincremental}
1. Create an R project (`.Rproj`)
2. Paste in the raw data (`raw` folder)
3. Write R scripts to analyze the data (`scripts` folder)
4. Reference helper functions within the R scripts (`R` folder)
5. Create a notebook to document the results (`.Rmd` file)
6. Create a simple script to run all the other scripts (`.R` file)
:::

## Final Thought

- Why did we do all this?
  - This is a self-contained, well-organized project that can be easily executed by others (*or your future self!*)

## CLASS ACTIVITY! {.center}

```{=html}
<style>
.small-text {
  font-size: 0.75em;
}

.tight-break {
  line-height: 0.5;
}

.reveal pre {
  margin-top: 10px;
  margin-bottom: 10px;
}

.reveal .column.pad-left {
  padding-left: 50px;
  box-sizing: border-box;
}
</style>
```